### 2013-03-16 doc/sosp13/intro.tex

Grappa leverages as much freely available and commodity infrastructure as
possible. We use unmodified Linux for the operating system and an
off-the-shelf user-mode infiniband device driver stack~\cite{OFED}. MPI is
used for process setup and tear down. GASnet~\cite{gasnet} is used as the
underlying mechanism for remote memory reads and writes using active message
invocations. To this commodity hardware and software mix Grappa adds three
main software components: (1) a \emph{lightweight tasking\/} layer that supports
a context switch in as few as 38ns and distributed global load balancing; (2)
a \emph{distributed shared memory\/} layer that supports normal access operations
such as \emph{read\/} and \emph{write\/} as well as synchronizing operations such as \emph{fetch-and-add\/}~\cite{fetchandadd}; and (3) a \emph{message aggregation\/} layer
that combines short messages to mitigate the aforementioned problem that
commodity networks are designed to achieve peak bandwidth only on large packet
sizes, yet irregular applications tend to fetch only a handful of bytes at a
time. As we will show later, Grappa can tolerate latencies way beyond that of
the network. Therefore, Grappa can afford to \emph{trade latency for
throughput\/}: by {\em increasing\/} latency in key components of the system we
are able to increase our effective random access memory bandwidth (by delaying and aggregating messages), our synchronization bandwidth (by delegating operations
to remote nodes), and our ability to improve load imbalance (work stealing
increases latency).

>>>>

Grappa leverages as much freely available and commodity infrastructure as
possible. We use unmodified Linux for the operating system and an
off-the-shelf user-mode infiniband device driver stack~\cite{OFED}. MPI is
used for process setup and tear down. GASnet~\cite{gasnet} is used as the
underlying mechanism for remote memory reads and writes using active message
invocations. To this commodity hardware and software mix Grappa adds three
main software components: (1) a \emph{lightweight tasking\/} layer that supports
a context switch in as few as 38ns and distributed global load balancing; (2)
a \emph{distributed shared memory\/} layer that supports normal access operations
such as \emph{read\/} and \emph{write\/} as well as synchronizing operations such as \emph{fetch-and-add\/}~\cite{fetchandadd}; and (3) a \emph{message aggregation\/} layer
that combines short messages to mitigate the aforementioned problem that
commodity networks are designed to achieve peak bandwidth only on large packet
sizes, yet irregular applications tend to fetch only a handful of bytes at a
time. As we will show later, Grappa can tolerate latencies way beyond that of
the network. Therefore, Grappa can afford to \emph{trade latency for
throughput\/}. By {\em increasing\/} latency in key components of the system we
are able to: increase effective random access memory bandwidth by delaying and aggregating messages; increase synchronization rate 
%this is kind of subtle -- do we just assert it or instead provide
%an inadequate explanation?  
%we delegate synchronization operations on data even when it resides on the
%local node so as to increase the rate at which we can perform
%read-modify-write operations in the event of a hot-spot, at the expense
%of increased synchronization latency as seen by tasks on that node. yuck.
by delegating atomic operations to gatekeeper cores, even when referencing node-local global data; and improve load balance by tolerating the delays incurred when work-stealing.

### 2013-03-16 doc/sosp13/memory.tex

\TODO{Talk about synchronous and asynchronous delegate operations. Talk about global completion object. Should we have a whole section on synchronization?}
\TODO{If we do want a section on GlobalCompletionEvent:
  \begin{itemize}
    \item many tasks and operations in the applications we've implemented do not need immediate consistency
    \item guaranteeing ordering requires blocking a task. this is unavoidable if the task relies on the returned value, or if one task is dependent on others
    \item however, much of the applications we've implemented (and other big-data things) are data-parallel, so tasks within a data-parallel loop can run in any order
    \item in addition, often these data-parallel tasks will update values that aren't used until the next data-parallel loop
    \item this allows us to free up workers/stacks (running out of stacks causes  latency-sensitivity because we can't continue until tasks complete)
    \item must guarantee all operations that were supposed to happen in the loop complete before continuing, including tasks and delegate operations
    \item \emph{motivating example:} write to array in one loop, use updated values in next loop (could also mention the ``PushBuffer'' example where we buffer queue appends because the order in the queue doesn't matter)
    \item GlobalCompletionEvent is complicated by task-stealing and the fact that delegate operations complete on a different core than they started, forcing us to use a ``cancelable barrier''-like approach
    \item interface to GlobalCompletionEvent:
      \begin{itemize}
        \item \texttt{gce.enroll(n)}: enroll n tasks/operations to be done, when goes from 0 to 1, leave cancelable barrier
        \item \texttt{gce.complete(n)}: complete tasks, when it goes to zero, enter global cancelable barrier
        \item \texttt{gce.wait()}: block until all cores have no work
      \end{itemize}
    \item to prevent early wakeups, stolen tasks must send a completion message back to the core they were started on
  \end{itemize}
}

>>>>

\TODO{Talk about synchronous and asynchronous delegate operations. Talk about global completion object. Should we have a whole section on synchronization?}
\TODO{If we do want a section on GlobalCompletionEvent:
  \begin{itemize}
    \item many tasks and operations in the applications we've implemented do not need immediate consistency
    \item guaranteeing ordering requires blocking a task. this is unavoidable if the task relies on the returned value, or if one task is dependent on others
    \item however, much of the applications we've implemented (and other big-data things) are data-parallel, so tasks within a data-parallel loop can run in any order
    \item in addition, often these data-parallel tasks will update values that aren't used until the next data-parallel loop
    \item this allows us to free up workers/stacks (running out of stacks causes  latency-sensitivity because we can't continue until tasks complete)
    \item must guarantee all operations that were supposed to happen in the loop complete before continuing, including tasks and delegate operations
    \item \emph{motivating example:} write to array in one loop, use updated values in next loop (could also mention the ``PushBuffer'' example where we buffer queue appends because the order in the queue doesn't matter)
    \item GlobalCompletionEvent is complicated by task-stealing and the fact that delegate operations complete on a different core than they started, forcing us to use a ``cancelable barrier''-like approach (possible citation: Ravichandran, et al. ``Work stealing for multi-core HPC Clusters'')
    \item interface to GlobalCompletionEvent:
      \begin{itemize}
        \item \texttt{gce.enroll(n)}: enroll n tasks/operations to be done, when goes from 0 to 1, leave cancelable barrier
        \item \texttt{gce.complete(n)}: complete tasks, when it goes to zero, enter global cancelable barrier
        \item \texttt{gce.wait()}: block until all cores have no work
      \end{itemize}
    \item to prevent early wakeups, stolen tasks must send a completion message back to the core they were started on
  \end{itemize}
}

### 2013-03-16 doc/sosp13/relatedwork.tex

\paragraph{Global memory} Grappa includes support for global memory namespace,
effectively a custom implementation of a software distributed shared memory
(DSM) system.

@@

\paragraph{Multithreading} Grappa uses multithreading to tolerate memory
latency. This is a well known technique. Hardware implementations include the
Tera MTA~\cite{tera:mta1}, Cray XMT~\cite{feo:xmt}, Simultaneous
multithreading~\cite{tullsen:smt}, MIT Alewife~\cite{agarwal:alewife},
Cyclops~\cite{almasi:cyclops}, and even GPUs~\cite{gpus}. As we describe in
this paper, Grappa implements its own software-based multithreading with a
lightweight user-mode task scheduler to multiplex \emph{thousands\/} of tasks
on a single processing core. The large number of tasks is required because of
the to tolerate very high inter-node communication latency. Grappa's task
library employs several optimizations: an extremely fast task switch, a small
task size, and judicious use of hardware prefetching to bring task state into
the cache long before that task is actually scheduled. The main difference between Grappa's support for lightweight threads and prior work such as QThread~\cite{qt} and Capriccio~\cite{capriccion} is the pervasive use of prefetching (with a large number of active tasks, not even their contexts fit in cache!).

>>>>

\paragraph{Global memory} Grappa supports a global memory namespace, presenting a programming abstraction similar to that of previous software distributed shared memory (DSM) systems but with differences attuned to irregular application demands.

@@

\paragraph{Multithreading} Grappa uses multithreading to tolerate memory
latency. This is a well known technique. Hardware implementations include the
Tera MTA~\cite{tera:mta1}, Cray XMT~\cite{feo:xmt}, Simultaneous
multithreading~\cite{tullsen:smt}, MIT Alewife~\cite{agarwal:alewife},
Cyclops~\cite{almasi:cyclops}, and even GPUs~\cite{gpus}. As we describe in
this paper, Grappa implements its own software-based multithreading with a
lightweight user-mode task scheduler to multiplex \emph{thousands\/} of tasks
on a single processing core. The large number of tasks is required because of
the to tolerate very high inter-node communication latency. Grappa's task
library employs several optimizations: an extremely fast task switch, a small
task size, and judicious use of hardware prefetching to bring task state into
the cache long before that task is actually scheduled. The main difference between Grappa's support for lightweight threads and prior work such as QThread~\cite{Wheeler08qthreads:an} and Capriccio~\cite{Behren03capriccio:scalable} is the pervasive use of prefetching (with a large number of active tasks, not even their contexts fit in cache!).

### 2013-03-16 doc/sosp13/tasks.tex

During execution, a task yields control of its core whenever it performs
a long-latency operation, allowing the processor to remain busy while
waiting for the operation to complete.  In addition,  a programmer can
direct scheduling explicitly via the Grappa API calls shown
in Figure~\ref{fig:scheduling}.
To minimize yield overhead, the Grappa scheduler operates entirely in
user-space and does little more than store register state of one task and load
that of another. Context switch times are as low as 38ns even when switching
amongst thousands of tasks.

>>>>

During execution, a task yields control of the processor core whenever it
performs a long-latency operation, allowing the processor to remain busy while
waiting for the operation to complete. In addition, a programmer can direct
scheduling explicitly via the Grappa API calls shown in
Figure~\ref{fig:scheduling}. To minimize yield overhead, the Grappa scheduler
operates entirely in user-space and does little more than store register state
of one task and load that of another. Context switch times are as low as 38ns
even when switching amongst thousands of tasks.
\TODO{We should expand on the cleverness of our scheduler?}
